---
title: "Lab 10"
subtitle: "ESS 330"
author: "Mckenna Cooper"
format: html
execute: 
  echo: true
---
## Libraries
```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)
library(gghighlight)
library(ggrepel)
library(knitr)
```

# Question 1

## 1.1 Define a Projection
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

## 1.2 Get USA Boundaries
```{r}
remotes::install_github("mikejohnson51/AOI")
library(AOI)
conus_states <- aoi_get(state = 'conus')
conus_states_eqdc <- st_transform(conus_states, crs = eqdc)
```

## 1.3 Get country boundaries for Mexico, the United States of America, and Canada
```{r}
countries <- aoi_get(country = c("USA", "MX", "CA"))

# Reproject to Equidistant Conic
countries_proj <- st_transform(countries, crs = eqdc)
```

## 1.4 Get city locations from the CSV file
```{r}
readr::read_csv("data/uscities.csv")

cities_raw <- read_csv("data/uscities.csv")

glimpse(cities_raw)
```
```{r}
cities_sf <- st_as_sf(cities_raw, coords = c("lng", "lat"), crs = 4326) 
```

```{r}

cities_proj <- st_transform(cities_sf, crs = eqdc)

cities_conus <- st_join(cities_proj, conus_states_eqdc, join = st_within, left = FALSE)

```

# Question 2

## 2.1 Distance to USA Border (coastline or national) (km)
```{r}
library(flextable)

usa_outline <- st_union(conus_states_eqdc)

usa_border <- st_cast(usa_outline, "MULTILINESTRING")

cities_conus$dist_to_us_border_km <- as.numeric(st_distance(cities_conus, usa_border)) / 1000

top5_cities <- cities_conus %>%
  arrange(desc(dist_to_us_border_km)) %>%
  slice(1:5) %>%
  select(city, state_name.x, dist_to_us_border_km) %>%
  st_drop_geometry()

top5_cities_flex <- flextable(top5_cities) %>%
  
  set_header_labels(
    city = "City",
    state_name.x = "State",
    dist_to_us_border_km = "Distance to Border (km)"
  ) %>%
  add_header_row(values = "Top 5 Farthest Cities from USA Border", colwidths = 3)

top5_cities_flex
```

## 2.2 Distance to States (km)
```{r}
states_lines <- st_cast(conus_states_eqdc, "MULTILINESTRING")

cities_dist_to_state_km <- apply(st_distance(cities_proj, states_lines), 1, min) / 1000

cities_proj$dist_to_state_km <- cities_dist_to_state_km

top5_cities_state <- cities_proj %>%
  arrange(desc(dist_to_state_km)) %>%
  select(city, state_name, dist_to_state_km) %>%
  head(5) %>%
  st_drop_geometry() 

top5_cities_state_flex <- flextable(top5_cities_state) %>%
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_state_km = "Distance to State Border (km)"
  ) %>%
  add_header_row(values = "Top 5 Farthest Cities from State Border", colwidths = 3)

top5_cities_state_flex
```

## 2.3 Distance to Mexico (km)
```{r}
mexico <- countries_proj %>%
  filter(sovereignt == "Mexico")

mexico_proj <- st_transform(mexico, st_crs(cities_proj))

cities_proj$dist_to_mexico_km <- st_distance(cities_proj, mexico_proj) / 1000

top5_cities_mexico <- cities_proj %>%
  select(-city_ascii, -state_id, -county_fips, -county_name, -population, 
         -density, -source, -military, -incorporated, -timezone, -ranking, 
         -zips, -id, -dist_to_state_km) %>%
  arrange(desc(dist_to_mexico_km)) %>%
  head(5) %>%
  st_drop_geometry()

top5_cities_mexico_flex <- flextable(top5_cities_mexico) %>%
  set_header_labels(
    city = "City",
    state_name = "State",
    dist_to_mexico_km = "Distance to Mexico (km)"
  ) %>%
  add_header_row(values = "Top 5 Farthest Cities from Mexico", colwidths = 3)


top5_cities_mexico_flex
```
**Note: I believe that the distance is in the correct units. I'm not sure why there is an [m] next to the distance values. **

## 2.4 Distance to Canada (km)
```{r}
# 1. Isolate Canada from the country objects
canada_proj <- countries_proj %>%
  filter(sovereignt == "Canada")

# 2. Calculate the distance from each city to the Canadian border (in kilometers)
cities_proj$dist_to_canada_km <- as.numeric(st_distance(cities_proj, canada_proj)) / 1000

# 3. Create a table for the five cities farthest from the Canadian border
top5_cities_canada <- cities_proj %>%
  select(-city_ascii, -state_id, -county_fips, -county_name, -population, 
         -density, -source, -military, -incorporated, -timezone, -ranking, 
         -zips, -id, -dist_to_state_km, -dist_to_mexico_km) %>%
  arrange(desc(dist_to_canada_km)) %>%
  head(5) %>%
  st_drop_geometry()  # Remove the geometry column for display in the table

# 4. Create the flextable with only the desired columns
top5_cities_canada_flex <- flextable(top5_cities_canada) %>%
  set_header_labels(
    name = "City",
    state_name = "State",
    dist_to_canada_km = "Distance to Canada (km)"
  )

# Display the flextable
top5_cities_canada_flex

```

# Question 3

## 3.1 Data
```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)

# Filter to 3 continents of interest (e.g., North America, South America, Europe)
continents <- countries_proj %>%
  filter(continent %in% c("North America", "South America", "Europe"))

# Get top 10 cities by population
top10_cities <- cities_conus %>%
  arrange(desc(population)) %>%
  slice_head(n = 10)

# Plot
ggplot() +
  geom_sf(data = continents, fill = "gray95", color = "gray70", lty = "dashed", size = 0.3) +
  geom_sf(data = conus_states_eqdc, fill = NA, color = "black", size = 0.4) +
  geom_sf(data = st_union(conus_states_eqdc), fill = NA, color = "red", size = 0.6, lty = "solid") +
  geom_sf(data = top10_cities, aes(size = population), color = "blue") +
  ggrepel::geom_label_repel(data = top10_cities,
                            aes(geometry = geometry, label = city),
                            stat = "sf_coordinates",
                            size = 3,
                            min.segment.length = 0) +
  theme_minimal() +
  labs(title = "Top 10 Largest U.S. Cities and Geographic Context",
       subtitle = "Includes 3 Continents, CONUS Outline, and State Boundaries") +
  theme(legend.position = "none")

```

## 3.2 City Distance from the Border
```{r}
# Reattach geometry to top5_cities
top5_cities_sf <- cities_conus %>%
  semi_join(top5_cities, by = c("city", "state_name.x", "dist_to_us_border_km"))

```

```{r}
library(ggplot2)
library(ggrepel)

ggplot() +
  # State boundaries
  geom_sf(data = conus_states_eqdc, fill = NA, color = "gray30", size = 0.3) +
  
  # All cities colored by distance
  geom_sf(data = cities_conus, aes(color = dist_to_us_border_km), size = 1) +
  
  # Highlight top 5 cities
  geom_sf(data = top5_cities_sf, color = "red", size = 2) +
  
  # Labels for top 5 cities
  ggrepel::geom_label_repel(
    data = top5_cities_sf,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    min.segment.length = 0
  ) +
  
  scale_color_viridis_c(option = "plasma", name = "Distance to Border (km)") +
  
  labs(
    title = "U.S. Cities Colored by Distance from National Border",
    subtitle = "Top 5 Farthest Cities Highlighted",
    caption = "Data source: cities_conus"
  ) +
  
  theme_minimal()

```

## 3.3 City Distance from Nearest State
```{r}
library(dplyr)
library(viridis)

# 5 farthest from state border
top5_state_cities <- cities_proj %>%
  arrange(desc(dist_to_state_km)) %>%
  slice_head(n = 5)

ggplot() +
  geom_sf(data = conus_states_eqdc, fill = NA, color = "black", size = 0.3) +
  geom_sf(data = cities_proj, aes(color = dist_to_state_km), size = 0.5) +
  scale_color_viridis(name = "Distance to State Border (km)", option = "viridis") +
  geom_sf(data = top5_state_cities, color = "blue", size = 2) +
  ggrepel::geom_label_repel(data = top5_state_cities, 
                            aes(label = city, geometry = geometry), 
                            stat = "sf_coordinates",
                            min.segment.length = 0) +
  theme_minimal() +
  labs(title = "Distance of USA Cities from Nearest State Border")

```


